A Review of Mortality Models

### 2020-09-08 mortality-models/draft.tex
Some more updates.

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\renewcommand{\vec}[1]{\mathbf{#1}}

\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behaviour of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

@@

In this article we shall review the development mathematical models to study
and forecast mortality in large enough populations of people.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behaviour of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

@@

Before we start reviewing the deterministic and stochastic mortality models
let us describe certain characteristic features of human mortality. Although
the life expectancy at birth has significantly improved in the post second
world war years the maximum life span seems to be pretty constant throughout 
history. A new born child faces a great risk of dying as it adjusts itself
to its new environment, warding off pathogens and fighting diseases. As time 
goes by the child starts developing immunity against many germs and its
chances of dying begin to drop. This drop continues until early adulthood 
when there is a sudden jump in mortality. It happens because of aggressive
behaviour in the late teens in both sexes and because of maternal mortality
in females. The spike in mortality eventually dwindles into a monotonic
increase in mortality as senescence catches up. We will see that there is 
a rich collection of mathematical models describing these characteristics
of human mortality. The basic structure of the mortality behaviour also
suggests that it might be profitable to fit a different model for different
stages of human life. The earliest models indeed did not cover childhood or
take into account the spike in mortality starting late teens. They were 
simple mathematical functions that fit a portion of the mortality curve that
was of most relevance to the insurance companies. Later models continued to
be deterministic in nature but had more number of parameters that could be
fitted for a given empirical data. However, it was not always possible to 
ascribe an unambiguous biological meaning to all parameters. Further, their
complexity was also controlled by the limitations of the computing technology
of their period.

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalisations in section 4. In section 5 we mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section 6. Starting
section 7 we turn our attention to stochastic models, starting with the
celebrated Lee-Carter model. An interesting class of models used across social
sciences is the Age-Period-Cohort models. We review a few of them relevant
to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
and finally we consider stochastic models with cohort effects in section 10.
